---
title: "Analysing Geodata in R - YoMos 2026"
author: "Yasemin Kukla & Jan Steen"
format: 
  revealjs: 
    embed-resources: true
    scrollable: true
    smaller: true
  html:     
    self-contained: false
    includes:
      in_header: rglwidget.html
editor: visual
footer: "YoMos Workshop 2026 - Analysing geodata in R"
code-fold: show
---

## Geodata

::: {style="font-size: 125%;"}
simplified definition:

-   Digital information, which is associated with a specific position on earth
:::

::: {.incremental .absolute bottom="200" left="0" style="font-size: 125%;"}

- Who did previous work with geodata?
- What kinds of geodata do you know?

:::

## Types of geodata 

::::::: columns 
:::: {.column width="50%" style="font-size: 125%;"}
Vector data

::: incremental
-   points, lines or polygons
-   e.g. plant observations, streets, political borders
-   shapefile, geopackage, geoJSON, ...
:::
::::

:::: {.column .fragment width="50%" style="font-size: 125%;"}
Raster data

::: incremental
-   grid filled with information
-   e.g. satellite images, corine land cover
-   .tif files
:::
::::
:::::::

## Using R for analysing geodata

::: {.incremental style="font-size: 125%;"}
-   open source platform
-   flexible (statistical) analysis and visualisation
-   automation and replication of workflows
-   easy sharing of code through git
-   various packages for special use cases
:::

## Basic R packages for geodata

-   "sf" for vector data
-   "terra" for raster data

::: notes
(maybe implement some links/images to those packages)
:::

```{r, echo = TRUE}
#install.packages("sf")
library(sf)

#install.packages("terra")
library(terra)
```

## First steps with vector data

Our example dataset 

::: {.incremental}
* OpenStreetMap (download.geofabrik.de/europe/germany/schleswig-holstein.html)
* administrative boundaries (https://geodaten.schleswig-holstein.de/gaialight-sh/_apps/dladownload/dl-vg_kkag.html)

:::

::: fragment
```{r read sf data, echo=TRUE,results='hide'}
landuse <- st_read("../data/vectordata/landuse.gpkg")
roads <- st_read("../data/vectordata/roads.gpkg")
buildings <- st_read("../data/vectordata/buildings.gpkg")
pois <- st_read("../data/vectordata/pois.gpkg")
```
:::

## Simple features in R

```{r showing sf objects, echo = TRUE}
str(landuse)
```

## Thematic queries

Where are benches around us?

```{r benches, echo = TRUE}
library(dplyr)
bench <- pois |> filter(fclass=="bench") 
```

::: fragment
```{r plot benches}
library(mapview)
mapview(bench)
```
:::

## Topological queries 1/2

Which benches are located in the heathland?

```{r benches in heathland, echo = TRUE}
heath <- landuse |> filter(fclass=="heath")

BenchInHeath <- st_intersects(bench,heath)
BenchInHeath <- bench[lengths(BenchInHeath) > 0,]
```

::: fragment
```{r plot heath benches}
mapview(heath, col.regions = "hotpink2")+mapview(BenchInHeath ,col.regions = "sienna")
```
:::

## Topological queries 2/2

Finding a place for a short break

```{r awi, echo=TRUE}
AWI <- buildings |> filter(name=="AWI Gästehaus")
```

::: fragment
```{r awi plot}
mapview(AWI)
```
:::

## Topological queries 2/2

Finding a place for a short break

```{r awi2, echo=TRUE}
AWI <- buildings |> filter(name=="AWI Gästehaus")

rest <- st_is_within_distance(bench, AWI, dist = 500)
```

::: fragment
```{r}
mapview(bench[lengths(rest) > 0,]) + mapview(AWI, col.regions="red")
```
:::

## Vector calculations 1/2

Mean area of Lister Heathlands

```{r heath area, echo = TRUE}
heath$area <- st_area(heath)
mean(heath$area)
```

## Vector calculations 2/2

Length of Streets in List
```{r streets, echo =TRUE}
roads$length <- st_length(roads)
mean(roads$length)

```

::: fragment
```{r}
mapview(roads)
```
:::



## First steps with raster data

## Raster-vector operations

## Map projections

::: {.incremental style="font-size: 125%;"}

- 2 dimensional image of the 3 dimensional earth
- maps can be either true in Area, shape, or distances
- hundreds of projections have been developed
- EPSG code defines the projection

:::

add a picture of a projection here


## Projections in R

::: {style="font-size: 125%;"}

- R does not project 'on the fly' 
- we have to check projections before performing calculations

:::

::: fragment
```{r projections sf, echo=TRUE}
bench
```

:::

## Reproject in R

Reproject Vector data
```{r reproject sf, eval=FALSE, echo=TRUE}
st_transform(x, st_crs(y))      ## use another sf object
st_transform(x, "EPSG:4326")    ## use an epsg code
```

::: fragment

Reproject Raster data
```{r reproject raster, eval=FALSE, echo=TRUE}
project(x, y)                   ## use another object
project(x, 4326)                ## use an epsg code
```

:::

::: fragment

adding missing information
```{r reproject raster2, eval=FALSE, echo=TRUE}
crs(x) <- "EPSG:4326"          
```

:::













